# WHITE WINE DATA SET ANALYSIS BY  `MOSTAFA TAREK` 
========================================================

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(psych)
library(corrplot)
library(memisc)
```

### load the data set 

first six rows 
```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
ww <- read.csv('./wineQualityWhites.csv')
ww <- dplyr::select(ww,-X)
head(ww)

```
more details using str and summary functions on the data set 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
str(ww)
summary(ww)


```

this data set contains 11 independant variables and one response variable (quality)
and 4898 observations 

# Univariate Plots Section

i will analysis the independent vars  

#### fixed acidity 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(fixed.acidity),data = ww)+
  geom_histogram(binwidth = 0.15)+
  coord_cartesian(xlim = c(3.8,10))
```

normal distribution 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$fixed.acidity)
```

summary of fixed.acidity


#### volatile.acidity

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(volatile.acidity),data = ww)+
  geom_histogram(binwidth = 0.01)+
  coord_cartesian(xlim = c(0.08,0.69))

```

almost normal distribution
```{r, echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$volatile.acidity)
```

summary of volatile acidity 

#### citric.acid

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(citric.acid),data = ww)+
  geom_histogram(binwidth = 0.01)+
  coord_cartesian(xlim = c(0,0.8))
```

normal distribution 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$citric.acid)

```

summary of citric acid 

#### residual.sugar

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(residual.sugar),data = ww)+
  geom_histogram(binwidth = 1)
```

right skewed 

transform x axis to log base ten to make it normal distribution so that it will be appropriate asssumption for linear models 

so : 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(residual.sugar),data = ww)+
  geom_histogram(bins = 50)+
  scale_x_continuous(trans = 'log10')
```

bimodal distribution 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$residual.sugar)

```

summary of residual sugar

#### chlorides

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(chlorides),data = ww)+
  geom_histogram(bins=200)+
  coord_cartesian(xlim=c(0,0.1))
```

normal distribution 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$chlorides)

```

summary of chlorides 

#### free.sulfur.dioxide

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(free.sulfur.dioxide),data = ww)+
  geom_histogram(bins = 70)+
  coord_cartesian(xlim = c(0,100))
```

normal distribution 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$free.sulfur.dioxide)

```

summary of free sulfur dioxide 

#### total.sulfur.dioxide

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(total.sulfur.dioxide),data = ww)+geom_histogram(bins = 100)
```

normal distribtution 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$total.sulfur.dioxide)

```

summary of total sulfur dioxide 

#### density

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(density),data = ww)+
  geom_histogram(bins=100)+
  coord_cartesian(xlim = c(0.9871,1.01))
```

normal distribution 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$density)

```

summary of density 

#### pH 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(pH),data = ww)+
  geom_histogram(bins = 50)
```

```{r,echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$pH)

```

summary of pH


#### sulphates

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(sulphates),data = ww)+
  geom_histogram(binwidth =0.01)+
  coord_cartesian(xlim = c(0.2,.85))

```

normal distibution 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$sulphates)

```

summary of sulphates 

#### alcohol

```{r,echo=FALSE, message=FALSE, warning=FALSE}
temp1 = ggplot(aes(alcohol),data = ww)+
  geom_histogram(binwidth = 0.5)

temp2 = ggplot(aes(alcohol),data = ww)+
  geom_histogram()+
  scale_x_log10()+
  xlab('log base 10 alcohol')

grid.arrange(temp1,temp2,ncol=1)

```

right skewed 

and the second plot (the log base 10 for x axis) is slightly different than the orgnal plot so 
the alcohol will be write skewed 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$alcohol)

```

summary of alcohol 

#### quality

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(quality),data = ww)+
  geom_histogram(binwidth = 1)
```

normal distribution 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
summary(ww$quality)

```

summary of quality

# Univariate Analysis

### What is the structure of your dataset?

The data set contains of 

- 11 independant vars 

- 1 response var (quality)

- 4898 observations of white wine quality 

### What is/are the main feature(s) of interest in your dataset?

- most of the ind vars are normal distributions which is usefull for the linear models assumptions 

- all of the ind var are almost normal distribution except
  - residual.sugar var is right skewed and then we take the log base 10 for x axis and it becomes modal distribution 

  - alcohol is right skewed even after scale_x_log base 10 transformation 


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The correlation between the ind vars and response var and among themselves and I will cover this in the bivariant section 

### Did you create any new variables from existing variables in the dataset?

no 

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

no I didn't since the data is TIDY, CLEAN and good QUALITY

# Bivariate Plots Section
i will plot the correlation coefficients of between the each variable using pairs.panels with pearson method  
```{r,echo=FALSE, message=FALSE, warning=FALSE}
pairs.panels(ww[,],method = "pearson")
```

- density and alcohol have a strong negative relationship which make the colinearity problem (r = -.75)

- density and residual sugar have a strong positive relationship which make the colinearity problem (r = +.85)

- density and total sulfur doxide  have a strong positive relationship which make the colinearity problem (r = +.5)

- free sulfur  dioxide and total sulfur dioxide  have a strong positive relationship which make the colinearity problem (r = +.62)

i will use the corrplot package and corrplot.mixed function to plot the correlations and its color 
```{r,echo=FALSE, message=FALSE, warning=FALSE}
temp <- cor(ww)
corrplot.mixed(temp, lower= 'number',lower.col = 'black', upper='color',tl.pos = 'lt',order = "AOE")
```

color indicates the correlation coefficients and it variate from blue (strong positive) to red (strong negative)




```{r,echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(alcohol,density),data = ww)+
  geom_jitter(alpha=1/10)+
  geom_smooth()+
  ggtitle('jitter between alcohol and density')

```

- density and alcohol have a strong negative relationship which make the colinearity problem (r = -.75)

```{r}
cor(ww$density,ww$alcohol)
```
correlation of density and alcohol


```{r,echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(residual.sugar,density),data = ww)+
    geom_jitter(alpha=1/10)+
  geom_smooth()+
  ggtitle('jitter residual sugar vs density ')
```

- density and residual sugar have a strong positive relationship which make the colinearity problem (r = +.85)

```{r}
cor(ww$residual.sugar, ww$density)
```
correlation of density and residual sugar 

```{r,echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(total.sulfur.dioxide,density),data = ww)+
  geom_jitter(alpha=1/10)+
  geom_smooth()+
  xlim(c(0,280))+
  ggtitle('jitter total sulfur dioxide vs density ')

```

- density and total sulfur doxide  have a strong positive relationship which make the colinearity problem (r = +.5)

```{r}
cat('correlation', cor(ww$total.sulfur.dioxide,ww$density))

```

correlation of total sulfur dioxide and desity 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(total.sulfur.dioxide, free.sulfur.dioxide),data = ww)+
  geom_jitter(alpha=1/10)+
  geom_smooth()+
  ggtitle(' jitter total vs free sulfur dioxide')

```

- free sulfur  dioxide and total sulfur dioxide  have a strong positive relationship which make the colinearity problem (r = +.62)

```{r}
cor(ww$total.sulfur.dioxide,ww$free.sulfur.dioxide)
```

correlation of total and free sulfur dioxide 



```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(factor(quality),alcohol),data = ww)+
  geom_boxplot()+
  stat_summary(geom='point', fun.y = mean,color='blue')+
  ggtitle('boxplots and the blue dots indicate the mean values ')

```

6 to 9 quality level have high alcohol level than 3 to 5 quality level 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(factor(quality),density),data = ww)+
  geom_boxplot()+
  stat_summary(geom='point', fun.y = mean,color='blue')+
  coord_cartesian(ylim = c(0.99,1))+
  ggtitle('boxplots and the blue dots indicate the mean values ')
```

high quality has low density levels 

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I examined which components can influence the wine quality in the dataset and how strong is the correlation between quality and other features. I found out that alcohol content and density can play a lot more important role in quality.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

- density and alcohol have a strong negative relationship which make the colinearity problem (r = -.75)

- density and residual sugar have a strong positive relationship which make the colinearity problem (r = +.85)

- density and total sulfur doxide  have a strong positive relationship which make the colinearity problem (r = +.5)

- free sulfur  dioxide and total sulfur dioxide  have a strong positive relationship which make the colinearity problem (r = +.62)


### What was the strongest relationship you found?

residual.sugar and  density

# Multivariate Plots Section


i will use the geom point to plot the relation between residual sugar and density with the color = quality as categorical variable \
and the smooth method is 'lm' 

```{r ,  echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots}
# the strongest relatonship residual.sugar and  density
ggplot(aes(residual.sugar,density,color=factor(quality)),data = ww)+
      geom_point(alpha = 0.5, size = 1)+
  geom_smooth(method = "lm", se = FALSE, alpha=1)  +
  scale_color_brewer(type='seq') +
  scale_y_continuous(limits = c(0.987, 1.0025)) +
  scale_x_continuous(limits = c(0, 20))+
  theme_dark()+
  ggtitle('density vs residual sugar dioxide with quality')

```

hight quality make the relation between density and residual sugar strong positive 

i will use the geom point to plot the relation between total and free sulfur dioxide with the color = quality as categorical variable \
and the smooth method is 'lm'

```{r  ,  echo=FALSE , message=FALSE, warning=FALSE}
ggplot(aes(total.sulfur.dioxide,free.sulfur.dioxide,color=factor(quality)),data = ww)+
  geom_point(alpha=1/3)+
  scale_color_brewer(type='seq')+
  coord_cartesian(xlim=c(0,280), ylim=c(0,100))+
  geom_smooth(method = 'lm',se=FALSE)+
  theme_dark()+
  ggtitle('total vs free sulfur dioxide with quality')
```

total sulfur dioxide and free sulfur dioxide have a week correlation coefficient in high quality (9)  

# linear model 

split the data to train test data sets with 80% and 20% respectively
then bulid the three model and evaluate the min max accuracy for each model 

```{r,echo=FALSE , message=FALSE, warning=FALSE}
trainingRowIndex <- sample(1:nrow(ww), 0.8*nrow(ww))  # row indices for training data
trainingData <- ww[trainingRowIndex, ]  # model training data
testData  <- ww[-trainingRowIndex, ]   # test data

m1 <- lm(quality~ density + alcohol,data = trainingData)
m2 <- update(m1, ~ .+ chlorides + volatile.acidity + fixed.acidity)
m3 <- update(m2, ~.+ free.sulfur.dioxide+ total.sulfur.dioxide + citric.acid+ residual.sugar+ sulphates+ pH )

mtable(m1,m2,m3)

distPred <- predict(m3, testData)  # predict distance

actuals_preds <- data.frame(cbind(actuals=testData$quality, predicteds=distPred))  # make actuals_predicteds dataframe.
correlation_accuracy <- cor(actuals_preds)  # 82.7%
head(actuals_preds)
min_max_accuracy <- mean(apply(actuals_preds, 1, min) / apply(actuals_preds, 1, max))  

mma3 <- min_max_accuracy

distPred <- predict(m2, testData)  # predict distance

actuals_preds <- data.frame(cbind(actuals=testData$quality, predicteds=distPred))  # make actuals_predicteds dataframe.
correlation_accuracy <- cor(actuals_preds)  # 82.7%
min_max_accuracy <- mean(apply(actuals_preds, 1, min) / apply(actuals_preds, 1, max))  
mma2 <- min_max_accuracy

distPred <- predict(m1, testData)  # predict distance

actuals_preds <- data.frame(cbind(actuals=testData$quality, predicteds=distPred))  # make actuals_predicteds dataframe.
correlation_accuracy <- cor(actuals_preds)  # 82.7%
min_max_accuracy <- mean(apply(actuals_preds, 1, min) / apply(actuals_preds, 1, max))  
mma1 <- min_max_accuracy

```
summary of the three models and the first six rows of the predictions using model3 

```{r}
mma1
mma2
mma3
```
accuracy of the three model 1 2 and 3 respectively 

so to reduce the calculation time we use model 1 because model 2 and model 3 didn't do so much in min max accuracy 
note:

m1 using features density , alcohol

m2 using features density , alcohol , chlorides ,  volatile.acidity , fixed.acidity

m3 using all features 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

residual sugar and density

### Were there any interesting or surprising interactions between features?

interaction between the residual sugar and density in high quality they are very correlated to one another 

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

srengths > min max accuracy is 90%

limitations > the output is continous not discrete because i used a linear reg model not classification model like logistic reg 


------

# Final Plots and Summary


### Plot One

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=factor(quality), y=alcohol/max(alcohol)),data=ww)+
  geom_boxplot()+
  stat_summary(geom='point', fun.y=mean, color='red',shape=4)+
  ggtitle('quality  and alcohol % box plots with means in reds')+
  ylab('alcohol %')+
  xlab('quality')
```


### Description One

in quality 6, 7, 8, and 9 the alcohol level is higher than the quality in levels 3, 4 and 5 

the red dots indicate the mean value of each quality level 


### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x=factor(quality), y=density/max(density)),data=ww)+
  geom_boxplot()+
  stat_summary(geom='point', fun.y=mean, color='red',shape=4)+
  coord_cartesian(ylim = c(0.95,.97))+
  ggtitle('quality and density % box plots with means in reds')+
  xlab('quality')+
  ylab('density%')

```

### Description Two

in quality 6, 7, 8, and 9 the density level is lower than the quality in levels 3, 4 and 5 

the red dots indicate the mean value of each quality level 

### Plot Three

```{r echo=FALSE, Plot_Three}
ggplot(aes(density/max(density),alcohol/max(alcohol),color=factor(quality)),data = ww)+
  geom_point(alpha=1/5)+
  geom_smooth(method='lm')+
  scale_color_brewer(type='seq')+
  theme_dark()+
  coord_cartesian(xlim = c(0.95,0.97),ylim = c(.5,1))+
  ggtitle('desity% vs alcohol% with quality represents as color ')+
  ylab('alcohol%')+
  xlab('density%')
```


### Description Three

the density and alcohol have an inverse relationship 

in the high quality the alcohol level for specific density value is larger than the alcohol level in lower quality 

in other words in the same density value, the more alcohol level we get the more quality the wine tend to be 
------

# Reflection

My main feature of interest was wine quality and I could find some variables which tend to influence this factor, like alcohol content and density. In the first part I made some simple hitograms about the main factors and other variables as well. In the part of bivariate analysis, I tried to get a better knowledge on the factors which influence wine quality. I was suprised that residual sugar, pH, citric acid and fixed acidity do not correlate with wine quality.

i could make some other interesting explorations in the future, like trying another models for fit the data well like non linear regression model 

or using classification models with output from 3 to 9 
 
The project work took me more time than I previously estimated 

It took time to choose the best corresponding plots and to detect the variables that influence quality. I had to examine each variable separatly while not forgetting the whole picture which I found really challenging and exciting.